\bfseries Convolution-based nonlinear functions on tensor fields and tensor images

### 2022-08-27 paper/main.tex
Updates from Overleaf

% text macros
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}

% typesetting adjustments
\makeatletter
\renewcommand\section{\@startsection {section}{1}{\z@}%
  {-3.25ex \@plus -1ex \@minus -.2ex}%
  {1.5ex \@plus .2ex}%
  {\raggedright\normalfont\large\bfseries}}
\makeatother
\setlength{\textwidth}{5.00in}
\setlength{\oddsidemargin}{3.25in}
\addtolength{\oddsidemargin}{-0.5\textwidth}
\setlength{\textheight}{9.40in}
\setlength{\topmargin}{-0.50in}
\setlength{\footskip}{0.25in}
\pagestyle{myheadings}
\markboth{foo}{Hogg, Blum-Smith, \& Villar / Geometric filters for geometric images}
\newcommand{\figurerule}{\rule[1ex]{\textwidth}{0.2pt}}
\linespread{1.08}
\frenchspacing\sloppy\sloppypar\raggedbottom

\title{\bfseries%
Convolution-based functions on tensor fields and tensor images}
\author{David W. Hogg, Ben Blum-Smith, and Soledad Villar}
\date{}

@@

The symmetries that suggest these rules are continuous symmetries.
But of course images are usually---and for our purposes---discrete grids of values.
This suggests that in addition to the continuous symmetries respected by the tensor objects in the image pixels there will be discrete symmetries for each geometric image taken as a whole.
We will define these discrete symmetry groups and use them to define a useful kind of group equivariance for functions of geometric images.
This equivariance, it turns out, is very easy to enforce, even for nonlinear functions of geometric images, provided that we compose our nonlinear functions from simple geometric operations, including our geometric generalization of convolution.

@@

Convolutions can be followed by Kronecker contractions, Levi-Civita contractions and index permutations.
The Kronecker delta ($\delta_{ij}$) is the object with two indices $ij$ such that it has the value $+1$ when the two indices have the same value ($i=j$), and $0$ otherwise.
Below we define the contraction of a $k$-$p$-tensor (with $k\geq 2$) with one application of the Kronecker delta, which returns a $(k-2)$-$p$-tensor.
The Levi-Civita symbol in $d\geq 2$ spatial dimensions is the object with $d$ indices (where $d$ is the spatial dimension) such that if the $d$ indices are not repeated and in an even-permutation order the value is $+1$ and if the $d$ indices are not repeated and in an odd-permutation order the value is $-1$, and it has the value $0$ in all other cases.
We define the contraction of a $k$-$p$-tensor (with $k\geq (d-1)$) with one application of the Levi-Civita symbol, which returns a $(k-d+2)$-$(p+1)$-tensor, because the contraction sums over $d$ pairs of indices, and it changes the parity from even to odd or odd to even.
For one example of these contractions, the 2-tensor formed by the outer product of 1-tensors $a$ and $b$ can be contracted with the Kronecker delta to give the standard dot product $a^\top b = [a]_i\,[b]_j\,\delta_{ij}$, which is a 0-tensor or scalar.
For another example, the same 2-tensor can be contracted with the Levi-Civita symbol to give the standard cross product
$[a\times b]_k = [a]_i\,[b]_j\,\epsilon_{ijk}$, which is a pseudovector.

@@

Introduce the discrete group. Why discrete?

@@

Prove thing about invariant filters.

We will now create linear convolution operators that are group-equivariant, for groups of reflections, rotations, and translations (on the cubical torus).
We will construct these equivariant convolution operators by an averaging over group operators, followed by a matrix factorization (singular-value decomposition).

@@

\begin{definition}[Action of $B_d$ on $k$-tensors and $k$-pseudotensors]
\begin{align}
\mbox{$k$-tensor:} ~~ g\cdot(v_1 \otimes \ldots \otimes v_k) &= (g\cdot v_1)\otimes\ldots\otimes(g\cdot v_k) \\
\mbox{$k$-pseudotensor:} ~~ g\cdot(v_1 \otimes \ldots \otimes v_k) &= \det(g)\,(g\cdot v_1)\otimes\ldots\otimes(g\cdot v_k) ~,
\end{align}
And extend linearly
\end{definition}

@@

@@

\subsection{Construction of equivariant geometric convolution filters}
In the above we defined nonlinear equivariant geometric operators acting on $d$-dimensional images on the $d$-torus.
These, in turn, are constructed from products of the outputs of equivariant geometric convolution filters.
Here we show how to find all such filters by a group-averaging technique.

Briefly, the idea is the following:
We start with a set of independent $k'$-tensor (or $k'$-pseudotensor) convolution filters that span the space of all possible geometric convolution filters at side length $M$ ($M$ odd) and dimension $d$.
There are typically $M^d\,d^{k'}$ independent filters in this set.
We average the filters over all group operators $g\in B_d$, where $B_d$ is the group of 90-degree rotations and reflections.
We perform a singular value decomposition of these group-averaged filters to find the $p$ independent orthogonal filters that span the space of equivariant $k'$-tensor (or $k'$-pseudotensor) convolution filters at side length $M$ in $d$ dimensions.

In detail, group averaging happens HOW? That is, what is the exact mathematical operation?

\section{Expressive power of geometric convolutions}\label{sec:universality}

@@

How do we make the invariant filters?

How do we make the linear functions?

What do they look like?

Etc.

An example?

\begin{figure}[tp]
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
\includegraphics[width=\textwidth]{notebooks/filter+0_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter+1_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-1_2_3.png}
\includegraphics[width=\textwidth]{notebooks/filter+2_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-2_2_3.png}
  \end{center}
\caption{All the filters for $d=2$, $m=1$. HOGG: notes!}
  \end{mdframed}
\end{figure}

\begin{figure}[tp]
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
\includegraphics[width=\textwidth]{notebooks/filter+0_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-0_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter+1_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-1_2_5.png}
  \end{center}
\caption{All the filters for $d=2$, $m=2$. HOGG: notes!}
  \end{mdframed}
\end{figure}

Give a table saying how many filters there are for each $d$ and each $M$ at each $k$ and each parity.

@@

\begin{figure}
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
    \includegraphics[width=\textwidth]{notebooks/monomials_1.png}
  \end{center}
    \caption{A scalar image $s$ and convolutions with four chosen filters.}
  \end{mdframed}
\end{figure}

@@

@@

Note that pseudo-$k$-tensor filters only exist for certain triples of $(d, m, k)$.

\paragraph{Acknowlegements:}
This project made use of Python, numpy, matplotlib, and cmastro.

\end{document}

>>>>

% text macros
\newcommand{\sectionname}{Section}
\newcommand{\secref}[1]{\sectionname~\ref{#1}}
\newcommand{\figref}[1]{\figurename~\ref{#1}}

% typesetting adjustments
\makeatletter
\renewcommand\section{\@startsection {section}{1}{\z@}%
  {-3.25ex \@plus -1ex \@minus -.2ex}%
  {1.5ex \@plus .2ex}%
  {\raggedright\normalfont\large\bfseries}}% key thing is: \raggedright!
\makeatother
\setlength{\textwidth}{5.00in}
\setlength{\oddsidemargin}{3.25in}
\addtolength{\oddsidemargin}{-0.5\textwidth}
\setlength{\textheight}{9.40in}
\setlength{\topmargin}{-0.50in}
\setlength{\footskip}{0.25in}
\pagestyle{myheadings}
\markboth{foo}{Hogg, Blum-Smith, \& Villar / Geometric filters for geometric images}
\newcommand{\figurerule}{\rule[1ex]{\textwidth}{0.2pt}}
\linespread{1.08}
\frenchspacing\sloppy\sloppypar\raggedbottom

\title{\bfseries%
Convolution-based nonlinear functions on tensor fields and tensor images}
\author{David W. Hogg, Ben Blum-Smith, and Soledad Villar}
\date{}

@@

The symmetries that suggest these rules are continuous symmetries.
But of course images are usually---and for our purposes---discrete grids of values.
This suggests that in addition to the continuous symmetries respected by the tensor objects in the image pixels there will be discrete symmetries for each geometric image taken as a whole.
We will define these discrete symmetry groups and use them to define a useful kind of group equivariance for functions of geometric images.
This equivariance, it turns out, is very easy to enforce, even for nonlinear functions of geometric images, provided that we compose our nonlinear functions from simple geometric operations, including our geometric generalization of convolution.
When we enforce this equivariance, the convolution filters that appear look very much like the differential operators that appear in discretizations of vector calculus, which is perhaps not surprising.

@@

Convolutions can be followed by Kronecker contractions, Levi-Civita contractions and index permutations.
The Kronecker delta ($\delta_{ij}$) is the object with two indices $ij$ such that it has the value $+1$ when the two indices have the same value ($i=j$), and $0$ otherwise.
Below we define the contraction of a $k$-$p$-tensor (with $k\geq 2$) with one application of the Kronecker delta, which returns a $(k-2)$-$p$-tensor.
The Levi-Civita symbol in $d\geq 2$ spatial dimensions is the object with $d$ indices (where $d$ is the spatial dimension) such that if the $d$ indices are not repeated and in an even-permutation order the value is $+1$ and if the $d$ indices are not repeated and in an odd-permutation order the value is $-1$, and it has the value $0$ in all other cases.
We define the contraction of a $k$-$p$-tensor (with $k\geq (d-1)$) with one application of the Levi-Civita symbol, which returns a $(k-d+2)$-$(p+1)$-tensor, because the contraction sums over $d$ pairs of indices, and it changes the parity from even to odd or odd to even.
For one example of these contractions, the 2-tensor formed by the outer product of 1-tensors $a$ and $b$ can be contracted with the Kronecker delta to give the standard dot product $a^\top b = [a]_i\,[b]_j\,\delta_{ij}$, which is a 0-tensor or scalar.
For another example, the same 2-tensor can (in $d=3$ dimensions) be contracted with the Levi-Civita symbol to give the standard cross product
$[a\times b]_k = [a]_i\,[b]_j\,\epsilon_{ijk}$, which is a pseudovector.

@@

Introduce the discrete group. Why discrete? SOLE: SEE TEXT BELOW FOR DEFINITIONS.

@@

Prove things about invariant filters and equivariant linear functions.

@@

\begin{definition}[Action of $B_d$ on $k$-$p$-tensors]
\begin{align}
g\cdot(v_1 \otimes \ldots \otimes v_k) &= \det(g)^p\,(g\cdot v_1)\otimes\ldots\otimes(g\cdot v_k)
\end{align}
And extend linearly
\end{definition}

@@

HOGG: Make a table and a figure showing all the group operators in $B_d$ with $d=2$.

@@

\paragraph{Numerical construction of invariant geometric convolution filters:}
In the above we defined equivariant geometric operators acting on $d$-dimensional images on the $d$-torus.
These, in turn, are constructed from products of the outputs of convolutions with invariant geometric filters.
Here we show how to find all such filters by a group-averaging technique.

Briefly, the idea is the following:
We start with a set of independent $k'$-tensor (or $k'$-pseudotensor) convolution filters that span the space of all possible geometric convolution filters at side length $M$ ($M$ odd) and dimension $d$.
There are typically $M^d\,d^{k'}$ independent filters in this set.
We average the filters over all group operators $g\in B_d$, where $B_d$ is the group of 90-degree rotations and reflections.
We perform a singular value decomposition of these group-averaged filters to find the $p$ independent orthogonal filters that span the space of equivariant $k'$-tensor (or $k'$-pseudotensor) convolution filters at side length $M=2\,m+1$ in $d$ dimensions.
The results of these operations are shown for $d=2$ and $m\in[1, 2]$ in \figref{fig:filters23} and \figref{fig:filters25}.
HOGG: Give a table saying how many filters there are for each dimension $d$ and each side length $M$ at each order $k$ and each parity $p$.
\begin{figure}[tp]
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
\includegraphics[width=\textwidth]{notebooks/filter+0_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter+1_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-1_2_3.png}
\includegraphics[width=\textwidth]{notebooks/filter+2_2_3.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-2_2_3.png}
  \end{center}
\caption{All the filters for $d=2$, $m=1$ ($M=3$). HOGG: notes!\label{fig:filters23}}
  \end{mdframed}
\end{figure}
\begin{figure}[tp]
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
\includegraphics[width=\textwidth]{notebooks/filter+0_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-0_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter+1_2_5.png}\\
\includegraphics[width=\textwidth]{notebooks/filter-1_2_5.png}
  \end{center}
\caption{All the filters for $d=2$, $m=2$ ($M=5$). HOGG: notes!\label{fig:filters25}}
  \end{mdframed}
\end{figure}

\section{Expressive power of nonlinear convolution-based functions}\label{sec:universality}

@@

How do we make the invariant filters? See \secref{sec:equivariant}.

How do we make the linear functions? See \secref{sec:convolution}.

@@

What do they look like? Like calculus operators! See figures.
\begin{figure}
  \begin{mdframed}
  \color{captiongray}
  \begin{center}
    \includegraphics[width=\textwidth]{notebooks/monomials_1.png}
  \end{center}
    \caption{A scalar image $s$ and convolutions with four chosen filters.}
  \end{mdframed}
\end{figure}

@@

It is interesting that the convolution operations look (in many cases) very much like discrete versions of vector calculus operations, like div, grad, curl, and so on. That's interesting, and potentially useful.

@@

Note that $k$-$1$-tensor (pseudo) filters only exist for certain triples of $(d, m, k)$.

\paragraph{Acknowlegements:}
It is a pleasure to thank Teresa Huang (JHU) and Kate Storey-Fisher (NYU) for valuable discussions.
This project made use of Python~3 \cite{python3}, numpy \cite{numpy}, matplotlib \cite{matplotlib}, and cmastro \cite{cmastro}.
All code used for making the data and figures in this paper are available at \url{https://github.com/davidwhogg/GeometricConvolutions}.

\bibliographystyle{plain}
\raggedright
\bibliography{ref}
\end{document}
